Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Various edits in Policy.
Includes adding one aspect from discussion documented in Google Doc. The one
other aspect that is missing is "Is sustainability designed for or an emergent
property". However, I think this best goes into the "What does sustainability
mean" cross-cutting issue (if it isn't there already).

Frameworks provide a scaffold for thinking; they help us understand
which elements are important and how they relate. They also help us
break down a problem into chunks that can be pursued by different
groups or perhaps disciplines.

A group of papers submitted to the workshop provide frameworks with
these intentions. The frameworks have substantial overlap but with
different emphases and extents.  Accordingly we present them starting
with the area of most overlap and proceed to show the different
emphases and the range of factors included in the frameworks.

% \todo{seems like definitions will be mentioned elsewhere in the
% paper?}
Each of the framework papers included a definition of sustainability,
with many overlaps.  In essence the papers related the question to
change over time, an idea well captured in the definition used by the
UK's Scientific Software Sustainability Institute: ``software you use
today will be available---and continue to be improved and
supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This
idea was also expressed as software that continues to serve its
users~\cite{Pierce_WSSSPE}.  See \S\ref{sec:cross-cutting} for more on
the discussion of definitions of sustainability in the workshop.

One area in which there was not complete overlap was whether the word
(and thus the effort called for by the WSSSPE workshop) involved
environmental sustainability. Of course the word sustainability has
strong connotations from consideration of environmental issues,
evoking some mention of the areas in which software interacts with
overall environmental resource usage, particularly energy efficiency.
The two papers in this area which mentioned this did so without
integrating that analysis into the question of software being around
long-term, suggesting that questions of environmental impact of
scientific software is a conceptually distinct area of inquiry.

One group of papers presented frameworks that were primarily about
characteristics of software artifacts, connecting with the long
discourse on software quality. This approach is realized in adjectives
that can be applied to pieces of software but might also extend to
describe software projects.  Thus Calero and
colleagues~\cite{Calero_WSSSPE} propose adding elements to the ISO
standards for measuring software quality: proposing an additional
dimension of quality they call perdurability or ability to ``last over
time'' made up of three characteristics: reliability, maintainability
and adaptability. This shows some overlap with the framework employed
by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of
``extensibility, interoperability, maintainability, portability,
reusability and scalability,'' anticipating the sorts of work that
people would need to do with a software artifact in the future, ``as
stakeholders requirements, technology and environments evolve and
change.'' Venters et al. argue that these choices are really made very
early because they relate to the architecture of the software and
involve trade offs that ought to be analyzed alongside other
attributes of the software artifact. Lenhardt et
al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
lifecycle to argue for the inclusion of metadata throughout a piece of
software's life (discussing, for example, how it has been built and
tested and what ``data in'' and ``data out'' has been considered). In
addition, their analogy suggests that the software lifecycle might add
a phase of ``preservation'' and draw on understandings of what that
involves from studies of data.  In sum, then, these frameworks focused
on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of
resources to accomplish the goals of sustainability. The elements of
these frameworks focused far more on the organization of a software
project than they did on characteristics of the artifact itself
(although it is certainly true that the adjectives discussed above
could be applied to a software project). For example Pierce et
al~\cite{Pierce_WSSSPE} focus on the way that the project is run,
particularly in terms of how those involved communicate and jointly
set priorities, a process they call governance. In particular they
argue that because sustainability relates to having ongoing resources,
governance must be open to receive diverse input (by occurring online,
asynchronously, and publicly) and thus have the potential to
``transform passive users into active contributors''. They argue that
the Apache Software Foundation's incubation process teaches this and
could be learned from by projects throughout scientific software.
Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance,
describing two modes: ``top-down'' and ``bottom-up'' governance.  They
place governance alongside questions of the artifact (``technical''),
and questions of the origins of resources, in terms of who is funding
the work surrounding a piece of software (``the political'') and the
manner in which resources come to the project initially (commercial,
open source, closed partnerships, grant funded) and long-term (all
four plus paid support).

Frameworks also addressed themselves to the context in which software
projects exist (thus moving upwards in abstraction, from the artifact
itself, to the organization of its production and, in these
frameworks, the shape of the space which an artifact or project exists
in). These frameworks take the form of contingency theories in that
they outline a different set of challenges and argue that different
project organizations (and presumably artifact attributes) are
necessary to persist longterm in spaces with particular
characteristics.  Katz and Proctor~\cite{Katz_WSSSPE} propose thinking
of this idea of space in terms of three axes: temporal (long or short
term needs), spatial (local or global use) and purpose (specific to
general).  They propose that different project organizations will be
needed in different locations and argue that we should concentrate
research to understand those connections. Weber et
al~\cite{Weber_WSSSPE} describe their spaces by analogy with natural
ecosystems as ``niches'' which sustain particular pieces of
software. They define ``a software niche as the set of technical
requirements, organizational conditions and cultural mores that
support its maintenance and use over time.'' They call for better
understanding and modeling of niches (as well as further exploration
of the usefulness of ecosystem metaphors).

@@

The policy panel and discussion addressed the question of recognition
of work on scientific software and linked that recognition to
questions of reward and thus motivation for particular kinds of work
on scientific software. In short, software work in science was seen to
be inadequately visible in ways that ``count'' within the reputation
system underlying science. In his paper for this workshop, Katz placed
software work along with other ``activities that facilitate science
but are not currently rewarded or
recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
to ``support all researchers in presenting meaningful impact evidence
in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
Kneply et al. argued that the visibility of software that supported a
piece of science ``can have detrimental effects on funding, future
library development, and even scientific
careers.''~\cite{Knepley_WSSSPE}.

@@

Papers and discussion concentrated in three areas: How ought software
work to be visible, what are the barriers to its visibility, and what
can be done to make it more visible?

Most of the papers in this area focused on visibility of software in
scientific papers, since scientific papers are the most widely
accepted documentation of achievement in science. It was noted that
there appear to be no widely accepted standards on how the use of
software towards a paper ought to be mentioned, that journals,
citation style guides and other guides to scientific conduct are vague
about how to describe software. To address this, papers spoke of a
need for a fixed identifier for software, either directly through a
mechanism such as a Digital Object
Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper
written to document the software (and perhaps its creation), a
``software paper''~\cite{Chue_Hong_WSSSPE}.

Another approach is to try to reduce the difficulty of citing software
for authors, acknowledging that authors are often working with
software that itself wraps other software and therefore hides that
software. Knepley et al~\cite{Knepley_WSSSPE} approach this by
proposing a mechanism by which the software itself, after it has run,
provides the user with a set of citations that are relevant to the
code actually run. They describe a prototype implementation whereby
the citations are embedded in libraries and reported along with the
results, via a commandline interface~\cite{Knepley_WSSSPE}. Discussion
highlighted the difficulty that attempting to acknowledge the
contributions of all pieces of dependent code within a paper faces the
difficulty of creating very long citation lists, straining the analogy
of code used to papers cited. Katz approaches this issue by proposing
a system of ``transitive credit,'' recording dependencies and relative
contributions outside particular papers, relieving authors from the
responsibility of acknowledging each and every dependency. Instead
authors would acknowledge the percentage contribution of the software
they used directly and an external system would then be able to
recursively allocate that credit to those who had provided
dependencies. Finally Priem and Piwowar argued that machine learning
techniques could examine the body of published literature and extract
mentions of software, coping with the multitude of informal ways in
which authors mention software they have used~\cite{Priem_WSSSPE}.
Discussion included turning the emphasis from asking users to improve
their citation of software contributions, to ask how projects
producing software might monitor the literature to improve their
ability to show impact. Michael McLennan described the approach taken
by the NanoHub project to scan the literature using keywords and names
of known users to discover papers that are likely to have used their
software and platform, then to have graduate students read each paper,
highlighting any mention and perhaps following up with the authors to
locate stories of impact.  Work since the workshop has described this
process at \textit{publications.wikia.com}.

Acknowledgement in science does not come only in publications, of
course. A key location for visibility might be in the grant funding
process, both in bio-sketches and in grant project reporting. Some
progress has been made here. Representatives of the NSF at the meeting
emphasized these opportunities and encouraged participants to take
advantage of them. Nonetheless, there was skepticism that peer review
panels would value these contributions in the same ways as
publications.

Priem and Piwowar argued for additionally looking beyond publications
and drawing on evidence of contribution, and impact, recorded in other
online systems, such as source code repositories like Github, both
code contributions, downloads of code and dependencies as well as
conversations about software (mailing lists, twitter and
beyond)~\cite{Priem_WSSSPE}. They argue for providing scholars with
flexible resources so that they can tell their own stories in the
manner most appropriate for them and their audiences, a principle of
the ``altmetrics'' approach.

\todo{convert this sketch to discussion of policy options}
\begin{verbatim}
        Opportunities for Policy interventions:
                Funding agencies
                Journals.  Success of data policies, growth of software policies.
                Promotion and Tenure. Follow lead of Science of Team Science in surveying policies at universities and other institutes.
\end{verbatim}
\todo{review PiratePad notes for other aspects of credit-giving discussed.}

@@

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructue development built-in,
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which
computing frequently takes place in a distributed cloud, the control
over digital resources is increasingly shifting from physical
infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
et al observe, unlike for physical IT infrastructure in data centers,
science communities have developed very little in the way of practices and
technology for API governance, refered to by Krintz et al as the
``combined policy, implementation, and deployment control''. Web APIs
can and do change, sometimes quite frequently. raising the need to
port dependent applications. The effort required for porting is
notoriously difficult to estimate, making it nearly impossible for IT
organizations to assess and thus properly manage the impact of API
changes. To address this, Krintz et al propose a mechanism that
evaluates the porting effort between two versions of a web-service API
in a formal and automated way. To analyze the porting effort, they
divide API changes into \emph{syntactic similarity}, the changes in
inputs and outputs, and into \emph{semantic similarity}, the changes
in the API's behavior. In initial tests, their method showed good
congruence with human developers in scoring porting effort, offering
the possibility that API governance can become as solid a part of
scientific IT management as data center infrastructure management is
today.

Many facets of engineering for software sustainability strongly depend
on the maturity level of the software. However, the maturity level of
a software project meant to be sustained changes tremendously over its
development life cycle, and the eventual maturity level is often
difficult to predict during initial development. Using software
security as their case sudy, Heiland et al~\cite{Heiland_WSSSPE}
discuss how Maturity Models can be used to formalize best practices
appropriate for the different levels of maturity that a software
project may go through over its lifecycle. Cybersecurity is also an
example of a sustainability-relevant aspect in software design that is
rarely given due diligence in science. In particular in industry,
cybersecurity best practices for different stages of life cycle and
maturity have been formalized as Software Security Maturity Models,
and are widely used, yet awareness of these among scientific software
development communities remains low. In providing a path to tightening
security practices as software matures, such models align with the
objective of providing implementation approaches that the scientific
community can actually follow.

@@

% note that paper summaries are available in tag jameshowison-paper-summaries: 
% https://github.com/danielskatz/WSSSPE/blob/jameshowison-paper-summaries/WSSSPE1_report/WSSSPE1_summary_report.tex

>>>>

The workshop submissions grouped under this section provide frameworks
for thinking about, researching and understanding which elements of sustainability
are important and how they relate. Although there is substantial
overlap between the frameworks, they have different emphases and
extents.  Each paper in this group included a definition of sustainability,
with many overlaps.  In essence the papers related the question to
change over time, an idea well captured in the definition used by the
UK's Scientific Software Sustainability Institute: ``software you use
today will be available---and continue to be improved and
supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This
idea was also expressed as software that continues to serve its
users~\cite{Pierce_WSSSPE}.  A more thorough discussion of definitions
of software sustainability is presented in \S\ref{sec:cross-cutting}.

One area in which there was not complete overlap was whether the word
(and thus the effort called for by the WSSSPE workshop) involved
environmental sustainability. Of course the word sustainability has
strong connotations from consideration of environmental issues,
evoking some mention of the areas in which software interacts with
overall environmental resource usage, particularly energy efficiency.
The two papers in this area which mentioned this\todo{which are these
  papers? Should cite here.} did so without
integrating that analysis into the question of software being around
long-term, suggesting that questions of environmental impact of
scientific software is a conceptually distinct area of inquiry.

One group of papers presented frameworks that were primarily about
characteristics of software artifacts, connecting with the long
discourse on software quality. This approach is realized in adjectives
that can be applied to pieces of software but might also extend to
describe software projects.  Thus Calero et al.~\cite{Calero_WSSSPE} propose adding elements to the ISO
standards for measuring software quality. Specifically, they propose an additional
dimension of quality they call perdurability or ability to ``last over
time'' made up of three characteristics: reliability, maintainability
and adaptability. This shows some overlap with the framework employed
by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of
``extensibility, interoperability, maintainability, portability,
reusability and scalability,'' anticipating the sorts of work that
people would need to do with a software artifact in the future, ``as
stakeholders requirements, technology and environments evolve and
change.'' Venters et al. argue that these choices are really made very
early because they relate to the architecture of the software and
involve trade offs that ought to be analyzed alongside other
attributes of the software artifact. Lenhardt et
al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
lifecycle to argue for the inclusion of metadata throughout a piece of
software's life (discussing, for example, how it has been built and
tested and what ``data in'' and ``data out'' has been considered). In
addition, their analogy suggests that the software lifecycle might add
a phase of ``preservation'' and draw on understandings of what that
involves from studies of data.  In sum, then, these frameworks focused
on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of
resources to accomplish the goals of sustainability. The elements of
these frameworks focused far more on the organization of a software
project than they did on characteristics of the artifact itself
(although it is certainly true that the adjectives discussed above
could be applied to a software project). For example Pierce et
al.~\cite{Pierce_WSSSPE} focus on the way that the project is run,
particularly in terms of how those involved communicate and jointly
set priorities, a process they call governance. In particular they
argue that because sustainability relates to having ongoing resources,
governance must be open to receive diverse input (by occurring online,
asynchronously, and publicly) and thus have the potential to
``transform passive users into active contributors''. They argue that
the Apache Software Foundation's incubation process teaches this and
could be learned from by projects throughout scientific software.
Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance,
describing two modes: ``top-down'' and ``bottom-up'' governance.  They
place governance alongside questions of the artifact (``technical''),
and questions of the origins of resources, in terms of who is funding
the work surrounding a piece of software (``the political'') and the
manner in which resources come to the project initially (commercial,
open source, closed partnerships, grant funded) and long-term (all
four plus paid support).

Frameworks also concerned themselves with the context in which software
projects exist, moving in abstraction from the software artifact
itself to the organization of its production and
the shape of the space in which an artifact or project exists.
These frameworks take the form of contingency theories in that
they outline a different set of challenges and argue that different
project organizations (and presumably artifact attributes) are
necessary to persist longterm in spaces with particular
characteristics.  Katz and Proctor~\cite{Katz_WSSSPE} propose thinking
of this idea of space in terms of three axes: temporal (long or short
term needs), spatial (local or global use) and purpose (specific to
general).  They propose that different project organizations will be
needed in different locations and argue that we should concentrate
research to understand those connections. Weber et
al.~\cite{Weber_WSSSPE} describe their spaces by analogy with natural
ecosystems as ``niches'' which sustain particular pieces of
software. They define ``a software niche as the set of technical
requirements, organizational conditions and cultural mores that
support its maintenance and use over time.'' They call for better
understanding and modeling of niches (as well as further exploration
of the usefulness of ecosystem metaphors).

@@

How work on scientific software is recognized and rewarded strongly influences the motivation for particular kinds of work
on scientific software. A recurring theme of the panel discussion was
that software work in science is
inadequately visible within the reputation
system underlying science; in other words it often doesn't ``count''. In his paper for this workshop, Katz placed
software work along with other ``activities that facilitate science
but are not currently rewarded or
recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
to ``support all researchers in presenting meaningful impact evidence
in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
Kneply et al. argued that the visibility of software that supported a
piece of science ``can have detrimental effects on funding, future
library development, and even scientific
careers.''~\cite{Knepley_WSSSPE}.

@@

Papers and discussion concentrated on three overarching questions: How ought software
work to be visible, what are the barriers to its visibility, and what
can be done to make it more visible?

Most of the papers in this area focused on visibility of software in
scientific papers, since scientific papers are the most widely
accepted documentation of achievement in science. It was noted that
there appear to be no widely accepted standards on how the use of
software towards a paper ought to be mentioned, and that journals,
citation style guides and other guides to scientific conduct are vague
about how to describe software. To address this, papers advocated the
need for a fixed identifier for software, either directly through a
mechanism such as a Digital Object
Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper
written to document the software (and perhaps its creation), a
``software paper''~\cite{Chue_Hong_WSSSPE}. However, as was pointed
out during the panel discussion, one of the problems with papers as
the cited product is that their author list is fixed in time, which
discourages potential contributors who are not on the original author
list from designing incremental improvements as integration work
rather than separate (and hence possibly rewritten) software
products~\cite{howison_incentives_2013}.

Another approach is to reduce the difficulty of citing all software
underlying a research paper. For example, scientists often work with
software that itself wraps other software, leading to attribution
stacking that can make it non-obvious or even difficult to determine
what attribitions would be appropriate. Knepley et al.~\cite{Knepley_WSSSPE} approach this by
proposing a mechanism by which the software itself, after it has run,
provides the user with a set of citations, according to the pieces of
code actually executed. They describe a prototype implementation whereby
the citations are embedded in libraries and reported along with the
results, via a commandline interface~\cite{Knepley_WSSSPE}. Discussion
highlighted the difficulty that attempting to acknowledge the
contributions of all pieces of dependent code within a paper faces the
difficulty of creating very long citation lists, straining the analogy
of code used to papers cited. Katz approaches this issue by proposing
a system of ``transitive credit,'' recording dependencies and relative
contributions outside particular papers, relieving authors from the
responsibility of acknowledging each and every dependency. Instead
authors would acknowledge the percentage contribution of the software
they used directly and an external system would then be able to
recursively allocate that credit to those who had provided
dependencies. Finally Priem and Piwowar argued that machine learning
techniques could examine the body of published literature and extract
mentions of software, coping with the multitude of informal ways in
which authors mention software they used~\cite{Priem_WSSSPE}.
A point raised in the panel discussion was that instead of asking users to improve
their software citation practices, one can also ask how software
projects can better monitor the literature to improve their
ability to show impact. For example, the NanoHub project scans the literature using keywords and names
of known users to discover papers that are likely to have used their
software and platform, and assigns graduate students to read each paper,
highlighting mentions of software use and sometimes following up with the authors to
identify stories for demonstrating impact.  A process for tracking
software-using publications with the goal of increasing impact visibility is now described at \url{http://publications.wikia.com}.

Potential visibility, and thus acknowledgement of scientific software
products is not restricted to publications. Another key location for
visibility is in the grant funding
process, and as emphasized by NSF representatives at the meeting,
recent changes to grant proposal and reporting formats now allow both
applicants and awardees to report and highlight software products as
much as publications. Nonetheless, whether peer review
panels would value these contributions in the same way as
publications remains to be seen.

Priem and Piwowar argued that assessing the impact of software work
requires looking beyond publications, including evidence of contribution and impact recorded in social
coding-oriented resources such as Github, and conversations about software
in issue trackers, mailing lists, twitter and
beyond~\cite{Priem_WSSSPE}. In keeping with a principle of
the ``altmetrics'' approach, they advocate that scholars should have
resources that empower them to tell their own stories in the
manner most appropriate for them and their audiences.

@@

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al.~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al.~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructue development built-in,
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

Krintz et al.~\cite{Krintz_WSSSPE} describe how in an era in which
computing frequently takes place in a distributed cloud, the control
over digital resources is increasingly shifting from physical
infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
et al. observe, unlike for physical IT infrastructure in data centers,
science communities have developed very little in the way of practices and
technology for API governance, refered to by Krintz et al. as the
``combined policy, implementation, and deployment control''. Web APIs
can and do change, sometimes quite frequently. raising the need to
port dependent applications. The effort required for porting is
notoriously difficult to estimate, making it nearly impossible for IT
organizations to assess and thus properly manage the impact of API
changes. To address this, Krintz et al. propose a mechanism that
evaluates the porting effort between two versions of a web-service API
in a formal and automated way. To analyze the porting effort, they
divide API changes into \emph{syntactic similarity}, the changes in
inputs and outputs, and into \emph{semantic similarity}, the changes
in the API's behavior. In initial tests, their method showed good
congruence with human developers in scoring porting effort, offering
the possibility that API governance can become as solid a part of
scientific IT management as data center infrastructure management is
today.

Many facets of engineering for software sustainability strongly depend
on the maturity level of the software. However, the maturity level of
a software project meant to be sustained changes tremendously over its
development life cycle, and the eventual maturity level is often
difficult to predict during initial development. Using software
security as their case sudy, Heiland et al.~\cite{Heiland_WSSSPE}
discuss how Maturity Models can be used to formalize best practices
appropriate for the different levels of maturity that a software
project may go through over its lifecycle. Cybersecurity is also an
example of a sustainability-relevant aspect in software design that is
rarely given due diligence in science. In particular in industry,
cybersecurity best practices for different stages of life cycle and
maturity have been formalized as Software Security Maturity Models,
and are widely used, yet awareness of these among scientific software
development communities remains low. In providing a path to tightening
security practices as software matures, such models align with the
objective of providing implementation approaches that the scientific
community can actually follow.

@@
